---
title: "Immigration, Gender, and Citizenship Wage Disparities"
subtitle: "Exploratory & Diagnostic Visualizations"
author: "David Rios"
date: "Spring 2025"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: show
editor: visual
---

# Overview

This document generates figures, non-parametric returns to education, industry representation metrics, and model diagnostic plots for my research paper on gender, citizenship, and wage disparities in the U.S. labor market. 

The data are a pooled cross-section of the 2012--2019 American Community Survey (ACS). Every model-based figure is generated from the paper's **preferred specification**, estimated identically to `02_Regressions_and_Stats.do`.

Household income is deliberately excluded. Race/ethnicity is included alongside the birthplace fixed effects: birthplace absorbs origin-country factors and immigrant selection, while race absorbs racial stratification within and across origin groups.

# Libraries

```{r}
#| message: false
#| warning: false

library(here)
library(tidyverse)
library(haven)
library(scales)
library(viridis)
library(mgcv)
library(hexbin)
library(fixest)
library(broom)
```

# Data Import & Cleaning

```{r}
#| message: false
#| warning: false

# here() resolves from the .Rproj at the repo root, so this works for anyone
# who clones the repo without editing a hardcoded path.
acs_data <- haven::read_dta(here::here("Data", "clean", "clean_data.dta"))

# Filter wages & create log wage
acs_data <- acs_data %>%
  filter(incwage_adj > 0) %>%
  mutate(ln_wage = log(incwage_adj))

# Recode Demographic & Citizenship Labels
acs_data <- acs_data %>%
  mutate(
    citizen_cat_clean = as.numeric(citizen_cat),
    citizen_label = case_when(
      citizen_cat_clean == 0 ~ "Native Citizen",
      citizen_cat_clean == 1 ~ "Naturalized Citizen",
      citizen_cat_clean == 2 ~ "Non-Citizen",
      TRUE ~ "Native Citizen"
    ),
    sex_label = ifelse(sex == 1, "Male", "Female"),
    citizen_sex_lbl = case_when(
      citizen_cat_clean == 2 & sex == 2 ~ "Non-Citizen Female",
      citizen_cat_clean != 2 & sex == 2 ~ "Citizen Female",
      citizen_cat_clean == 2 & sex == 1 ~ "Non-Citizen Male",
      citizen_cat_clean != 2 & sex == 1 ~ "Citizen Male"
    )
  )

# Education: grouped attainment 
acs_data <- acs_data %>%
  mutate(
    educ_years = case_when(
      educ == 0         ~ 0,   # N/A or no schooling
      educ == 1         ~ 2,   # Nursery to Grade 4
      educ == 2         ~ 6,   # Grade 5-8
      educ == 3         ~ 9,   # Grade 9
      educ == 4         ~ 10,  # Grade 10
      educ == 5         ~ 11,  # Grade 11
      educ == 6         ~ 12,  # Grade 12 (HS Diploma)
      educ == 7         ~ 13,  # 1 yr college
      educ == 8         ~ 14,  # 2 yrs college
      educ == 9         ~ 15,  # 3 yrs college
      educ == 10        ~ 16,  # 4 yrs college (Bachelor's)
      educ == 11        ~ 18,  # 5+ yrs college (Graduate degree)
      TRUE              ~ NA_real_
    ),
    educ_grouped = case_when(
      educ <= 6              ~ "HS or Less",
      educ %in% c(7, 8, 9)   ~ "Some College",
      educ %in% c(10, 11)    ~ "Bachelor's or Higher"
    ),
    educ_grouped = factor(educ_grouped,
                          levels = c("HS or Less", "Some College", "Bachelor's or Higher"))
  )

# English Fluency & Children Groups
acs_data <- acs_data %>%
  mutate(
    eng_fluency = case_when(
      speakeng %in% c(3, 4) ~ "Fluent",
      speakeng %in% c(2, 5) ~ "Intermediate",
      speakeng %in% c(1, 6) ~ "Limited"
    ),
    eng_fluency = factor(eng_fluency, levels = c("Fluent", "Intermediate", "Limited")),
    nchild_grouped = case_when(
      nchild == 0            ~ "None",
      nchild == 1            ~ "One",
      nchild == 2            ~ "Two",
      nchild %in% c(3, 4)    ~ "Three-Four",
      nchild >= 5            ~ "Five+"
    ),
    nchild_grouped = factor(nchild_grouped,
                            levels = c("None", "One", "Two", "Three-Four", "Five+"))
  )

# Income Groups 
acs_data <- acs_data %>%
  mutate(
    income_tertile = ntile(incwage_adj, 3),
    income_group = case_when(
      income_tertile == 1 ~ "Low",
      income_tertile == 2 ~ "Medium",
      income_tertile == 3 ~ "High"
    ),
    income_group = factor(income_group, levels = c("High", "Medium", "Low"))
  )

# Industry & Birthplace
acs_data <- acs_data %>%
  mutate(
    industry_name   = as.character(haven::as_factor(industry_encoded)),
    birthplace_label = haven::as_factor(birthplace)
  )

# Compute Industry Ratio Metrics
ratio_data_detailed <- acs_data %>%
  filter(!is.na(industry_name)) %>%
  group_by(industry_name, citizen_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = citizen_label,
    values_from = count,
    values_fill = 0
  ) %>%
  mutate(
    ratio_nat = `Naturalized Citizen` / `Native Citizen`,
    ratio_non = `Non-Citizen` / `Native Citizen`
  ) %>%
  arrange(desc(ratio_nat + ratio_non))

ratio_long <- ratio_data_detailed %>%
  select(industry_name, ratio_nat, ratio_non) %>%
  pivot_longer(
    cols = starts_with("ratio"),
    names_to = "Group",
    values_to = "Ratio"
  ) %>%
  mutate(
    Group = recode(
      Group,
      ratio_nat = "Naturalized Citizen",
      ratio_non = "Non-Citizen"
    )
  )

# Center Age at the (unweighted) sample mean
mean_age <- mean(acs_data$age, na.rm = TRUE)
acs_data <- acs_data %>%
  mutate(
    age_c    = age - mean_age,
    age_c_sq = age_c^2
  )

# Preferred Specification 
full_fe_model <- feols(
  ln_wage ~ factor(citizen_cat_clean) * factor(sex) +
    age_c + age_c_sq +
    educ_grouped +
    eng_fluency +
    nchild_grouped +
    factor(rachsing) |
    year + industry_encoded + puma_state + birthplace,
  cluster = ~puma_state,
  data = acs_data
)

etable(full_fe_model, digits = 4)

# Extract Fitted Values and Residuals for Diagnostics
model_diagnostics <- data.frame(
  fitted_values = fitted(full_fe_model),
  residuals     = residuals(full_fe_model)
)
```


# Figures

## Figure 1: Log Wage Transformation & Distributional Symmetry

```{r}
#| fig-width: 8
#| fig-height: 4.5

fig1 <- acs_data %>%
  select(incwage_adj, ln_wage) %>%
  pivot_longer(cols = everything(), names_to = "metric", values_to = "value") %>%
  mutate(metric = ifelse(metric == "incwage_adj", "Raw Annual Wages ($2015)", "Log Annual Wages")) %>%
  ggplot(aes(x = value, fill = metric)) +
  geom_density(alpha = 0.6, show.legend = FALSE) +
  facet_wrap(~metric, scales = "free") +
  scale_fill_manual(values = c("Log Annual Wages" = "#21618c", "Raw Annual Wages ($2015)" = "#e74c3c")) +
  labs(
    title = "Distributional Transformation of Wages",
    subtitle = "Demonstrating Skewness Reduction via Natural Log Transformation",
    x = "Value",
    y = "Density"
  ) +
  theme_minimal(base_size = 11)

fig1
```



## Figure 2: Foreign-Born Demographic Diversity by Birthplace

```{r}
#| fig-width: 8
#| fig-height: 5

birthplace_counts <- acs_data %>%
  filter(!birthplace_label %in% c("U.S. & Territories", "Other/Unknown", "N/A")) %>%
  count(birthplace_label) %>%
  arrange(desc(n))

fig2 <- ggplot(birthplace_counts, aes(x = reorder(birthplace_label, -n), y = n)) +
  geom_bar(stat = "identity", fill = "#21618c") +
  geom_text(aes(label = comma(n)), vjust = -0.3, size = 3) +
  labs(
    title = "Demographic Diversity by Birthplace (Foreign-Born Only)",
    x = "Birthplace Region",
    y = "Count of Individuals") +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.1))) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9))

fig2
```



## Figure 3: Industry Concentration by Citizenship Status

```{r}
#| fig-width: 9
#| fig-height: 5

fig3 <- ggplot(ratio_long, aes(x = reorder(industry_name, Ratio), y = Ratio, fill = Group)) +
  geom_col(position = "stack", width = 0.8) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.05))) +
  scale_fill_viridis_d(option = "G", begin = 0.2, end = 0.8) +
  labs(
    title = "Foreign-Born Worker Concentration by Industry and Citizenship Status",
    subtitle = "Ratio of Foreign-Born Workers relative to Native-Born Citizen Workers (ACS 2012-2019)",
    x = "Industry Sector",
    y = "Ratio to Native-Born Citizen Workers (%)",
    fill = "Citizenship Status"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

fig3
```


## Figure 4: Intersectional Income Group Distribution

```{r}
#| fig-width: 10
#| fig-height: 6

educ_income_intersectional <- acs_data %>%
  filter(!is.na(educ_grouped)) %>%
  group_by(educ_grouped, income_group, citizen_label, sex_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(educ_grouped, citizen_label, sex_label) %>%
  mutate(percent = count / sum(count) * 100)

fig4 <- ggplot(educ_income_intersectional, aes(x = educ_grouped, y = percent, fill = income_group)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_viridis_d(name = "Income Tertile", option = "D") +
  labs(
    title = "Income Group Distribution by Gender, Citizenship, and Education",
    x = "Educational Attainment Group",
    y = "Percentage within Group"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 8),
    strip.text = element_text(size = 10, face = "bold")
  ) +
  facet_grid(sex_label ~ citizen_label)

fig4
```



## Figure 5: Non-Parametric Returns to Education (GAM Fit)

```{r}
#| fig-width: 9
#| fig-height: 5.5
#| warning: false

fig5 <- ggplot(acs_data, aes(x = educ_years, y = ln_wage, color = citizen_sex_lbl, fill = citizen_sex_lbl)) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = TRUE, alpha = 0.12, linewidth = 1.1) +
  scale_color_viridis_d(option = "plasma", name = "Demographic Group") +
  scale_fill_viridis_d(option = "plasma", name = "Demographic Group") +
  scale_x_continuous(breaks = seq(0, 18, by = 2)) +
  labs(
    title = "Non-Parametric Returns to Education by Citizenship & Gender",
    subtitle = "Pooled ACS 2012-2019 (GAM Smoothed Fit)",
    x = "Years of Formal Education",
    y = "Log Wage (2015 Constant USD)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

fig5
```


## Figure 6: Model Diagnostic: Residuals vs. Fitted Log Wages

```{r}
#| fig-width: 8.5
#| fig-height: 5

fig6 <- ggplot(model_diagnostics, aes(x = fitted_values, y = residuals)) +
  geom_hex(bins = 60) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#e74c3c", linewidth = 0.8) +
  scale_fill_viridis_c(option = "magma", trans = "log10", name = "Observation Count\n(Log Scale)") +
  labs(
    title = "Model Diagnostic: Residuals vs. Fitted Log Wages",
    subtitle = paste0("Preferred Specification, Hexagonal Binning (N = ",
                      comma(nrow(model_diagnostics)), ")"),
    x = "Fitted Log Wages",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 11)

fig6
```

## Figure 7: Coefficient Estimates with Clustered 95% CIs

```{r}
#| fig-width: 9
#| fig-height: 6

model_coefs <- tidy(full_fe_model, conf.int = TRUE)

model_coefs_clean <- model_coefs %>%
  filter(!term %in% c("(Intercept)", "age_c", "age_c_sq")) %>%
  mutate(
    term_clean = case_when(
      term == "factor(citizen_cat_clean)1" ~ "Naturalized Citizen",
      term == "factor(citizen_cat_clean)2" ~ "Non-Citizen",
      term == "factor(sex)2" ~ "Female",
      term == "educ_groupedSome College" ~ "Some College",
      term == "educ_groupedBachelor's or Higher" ~ "Bachelor's or Higher",
      term == "eng_fluencyIntermediate" ~ "English: Intermediate",
      term == "eng_fluencyLimited" ~ "English: Limited",
      term == "nchild_groupedOne" ~ "Children: One",
      term == "nchild_groupedTwo" ~ "Children: Two",
      term == "nchild_groupedThree-Four" ~ "Children: Three-Four",
      term == "nchild_groupedFive+" ~ "Children: Five+",
      term == "factor(rachsing)2" ~ "Race: Black",
      term == "factor(rachsing)3" ~ "Race: Amer. Indian/AK Native",
      term == "factor(rachsing)4" ~ "Race: Asian/Pacific Islander",
      term == "factor(rachsing)5" ~ "Race: Hispanic/Latino",
      term == "factor(citizen_cat_clean)1:factor(sex)2" ~ "Naturalized x Female",
      term == "factor(citizen_cat_clean)2:factor(sex)2" ~ "Non-Citizen x Female",
      TRUE ~ term
    )
  )

# Plot vertical coefficient plot
fig7 <- ggplot(model_coefs_clean, aes(x = estimate, y = reorder(term_clean, estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#e74c3c", linewidth = 0.8) +
  geom_pointrange(
    aes(xmin = conf.low, xmax = conf.high),
    color = "#21618c",
    size = 0.6,
    fatten = 2.5
  ) +
  labs(
    title = "Preferred Specification: Coefficient Estimates with Clustered 95% CIs",
    subtitle = "Year, Industry, Birthplace, and PUMA-State FE; SEs Clustered on PUMA-State",
    x = "Log Wage Impact (Estimate & 95% CI)",
    y = "Demographic & Human Capital Variable"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 9.5, color = "grey30"),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 9, color = "black")
  )

fig7
```


## Figure 8: Spatial Coverage of the PUMA-State Cluster Sample

Official Census Bureau TIGER/Line cartographic boundaries via `tigris`.

```{r}
#| fig-width: 9
#| fig-height: 5.5
#| message: false
#| warning: false

library(tigris)
library(sf)
options(tigris_use_cache = TRUE)

sampled_pumas <- acs_data %>%
  distinct(statefip, puma) %>%
  mutate(GEOID10 = sprintf("%02d%05d", as.integer(statefip), as.integer(puma)))

cat("Sampled PUMA-state geographies:", nrow(sampled_pumas), "\n")

# 2010-definition PUMA boundaries, all states
pumas_all <- tryCatch(
  pumas(cb = TRUE, year = 2019),
  error = function(e) {
    purrr::map_dfr(unique(fips_codes$state_code[as.integer(fips_codes$state_code) <= 72]),
                   ~ tryCatch(pumas(state = .x, cb = TRUE, year = 2019),
                              error = function(e) NULL))
  }
)

pumas_all <- pumas_all %>%
  # Drop island-territory PUMAs (AS/GU/MP/VI)
  filter(!STATEFP10 %in% c("60", "66", "69", "78")) %>%
  shift_geometry() %>%  # rescale AK/HI/PR into a compact national frame
  mutate(sampled = factor(GEOID10 %in% sampled_pumas$GEOID10,
                          levels = c(TRUE, FALSE),
                          labels = c("Sampled (118 PUMAs)", "Not sampled")))

states_all <- states(cb = TRUE, year = 2019) %>%
  filter(!STUSPS %in% c("AS", "GU", "MP", "VI")) %>%
  shift_geometry()

fig8 <- ggplot() +
  geom_sf(data = pumas_all, aes(fill = sampled), color = NA) +
  geom_sf(data = states_all, fill = NA, color = "grey40", linewidth = 0.2) +
  scale_fill_manual(values = c("Sampled (118 PUMAs)" = "#21618c",
                               "Not sampled" = "grey90"),
                    name = NULL) +
  labs(
    title = "Geographic Coverage of the 5% PUMA-State Cluster Sample",
    subtitle = "2010-Definition PUMAs; Sampled Geographies Shaded (ACS 2012-2019)"
  ) +
  theme_void(base_size = 11) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 12, face = "bold"),
    plot.subtitle = element_text(size = 9.5, color = "grey30")
  )

fig8
```


# Saving Images

```{r}
dir.create(here("Output", "figures"), recursive = TRUE, showWarnings = FALSE)

save_fig <- function(filename, plot_obj, w, h) {
  ggsave(
    filename = here("Output", "figures", filename),
    plot = plot_obj,
    width = w, height = h,
    dpi = 300, bg = "white"
  )
}

save_fig("Figure_1_Log_Wages.png",       fig1, 8,   4.5)
save_fig("Figure_2_Birthplace.png",      fig2, 8,   5)
save_fig("Figure_3_Industry.png",        fig3, 9,   5)
save_fig("Figure_4_Income.png",          fig4, 10,  6)
save_fig("Figure_5_Returns_to_Educ.png", fig5, 9,   5.5)
save_fig("Figure_6_Diagnostics.png",     fig6, 8.5, 5)
save_fig("Figure_7_Coefficients.png",    fig7, 9,   6)
save_fig("Figure_8_PUMA_Map.png",        fig8, 9,   5.5)

# PDF versions used directly by Paper/paper.tex
save_pdf <- function(filename, plot_obj, w, h) {
  ggsave(
    filename = here("Paper", filename),
    plot = plot_obj,
    width = w, height = h, bg = "white"
  )
}

save_pdf("fig1_industry_ratio.pdf", fig3, 9,   5)
save_pdf("fig2_birthplace.pdf",     fig2, 8,   5)
save_pdf("fig3_income_dist.pdf",    fig4, 10,  6)
save_pdf("fig4_returns_educ.pdf",   fig5, 9,   5.5)
save_pdf("fig5_coefplot.pdf",       fig7, 9,   6)
save_pdf("fig6_log_wages.pdf",      fig1, 8,   4.5)
save_pdf("fig7_puma_map.pdf",       fig8, 9,   5.5)
```
