---
title: "Immigration, Gender, and Citizenship Wage Disparities Revisited"
author: "David Rios"
date: "Spring 2025"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: show
---

# Overview

This document generates figures, returns to education, industry representation metrics, and model diagnostic plots for my research paper on gender, citizenship, and wage disparities in the U.S. labor market.

The data are a pooled cross-section of the 2012--2019 American Community Survey (ACS). Every model-based figure is generated from the paper's **preferred specification**, estimated identically to `02_Regressions_and_Stats.do`.

Household income is deliberately excluded. Race/ethnicity is included alongside the birthplace fixed effects: birthplace absorbs origin-country factors and immigrant selection, while race absorbs racial stratification within and across origin groups.

# Libraries

```{r}
#| message: false
#| warning: false

library(here)
library(tidyverse)
library(haven)
library(scales)
library(viridis)
library(fixest)
library(broom)
```

# Data Import & Cleaning

```{r}
#| message: false
#| warning: false

here::i_am("Code/03_Graphs.qmd")
acs_data <- haven::read_dta("/Users/davidrios/Econometrics_Final_Paper/Data/clean/clean_data.dta")

# Filter wages & create log wage
acs_data <- acs_data %>%
  filter(incwage_adj > 0) %>%
  mutate(ln_wage = log(incwage_adj))

# Recode Demographic & Citizenship Labels
acs_data <- acs_data %>%
  mutate(
    citizen_cat_clean = as.numeric(citizen_cat),
    citizen_label = case_when(
      citizen_cat_clean == 0 ~ "Native Citizen",
      citizen_cat_clean == 1 ~ "Naturalized Citizen",
      citizen_cat_clean == 2 ~ "Non-Citizen",
      TRUE ~ "Native Citizen"
    ),
    sex_label = ifelse(sex == 1, "Male", "Female"),
    citizen_sex_lbl = case_when(
      citizen_cat_clean == 2 & sex == 2 ~ "Non-Citizen Female",
      citizen_cat_clean != 2 & sex == 2 ~ "Citizen Female",
      citizen_cat_clean == 2 & sex == 1 ~ "Non-Citizen Male",
      citizen_cat_clean != 2 & sex == 1 ~ "Citizen Male"
    )
  )

# Education: grouped attainment 
acs_data <- acs_data %>%
  mutate(
    educ_years = case_when(
      educ == 0         ~ 0,   # N/A or no schooling
      educ == 1         ~ 2,   # Nursery to Grade 4
      educ == 2         ~ 6,   # Grade 5-8
      educ == 3         ~ 9,   # Grade 9
      educ == 4         ~ 10,  # Grade 10
      educ == 5         ~ 11,  # Grade 11
      educ == 6         ~ 12,  # Grade 12 (HS Diploma)
      educ == 7         ~ 13,  # 1 yr college
      educ == 8         ~ 14,  # 2 yrs college
      educ == 9         ~ 15,  # 3 yrs college
      educ == 10        ~ 16,  # 4 yrs college (Bachelor's)
      educ == 11        ~ 18,  # 5+ yrs college (Graduate degree)
      TRUE              ~ NA_real_
    ),
    educ_grouped = case_when(
      educ <= 6              ~ "HS or Less",
      educ %in% c(7, 8, 9)   ~ "Some College",
      educ %in% c(10, 11)    ~ "Bachelor's or Higher"
    ),
    educ_grouped = factor(educ_grouped,
                          levels = c("HS or Less", "Some College", "Bachelor's or Higher"))
  )

# English Fluency & Children Groups
acs_data <- acs_data %>%
  mutate(
    eng_fluency = case_when(
      speakeng %in% c(3, 4) ~ "Fluent",
      speakeng %in% c(2, 5) ~ "Intermediate",
      speakeng %in% c(1, 6) ~ "Limited"
    ),
    eng_fluency = factor(eng_fluency, levels = c("Fluent", "Intermediate", "Limited")),
    nchild_grouped = case_when(
      nchild == 0            ~ "None",
      nchild == 1            ~ "One",
      nchild == 2            ~ "Two",
      nchild %in% c(3, 4)    ~ "Three-Four",
      nchild >= 5            ~ "Five+"
    ),
    nchild_grouped = factor(nchild_grouped,
                            levels = c("None", "One", "Two", "Three-Four", "Five+"))
  )

# Income Groups 
acs_data <- acs_data %>%
  mutate(
    income_tertile = ntile(incwage_adj, 3),
    income_group = case_when(
      income_tertile == 1 ~ "Low",
      income_tertile == 2 ~ "Medium",
      income_tertile == 3 ~ "High"
    ),
    income_group = factor(income_group, levels = c("High", "Medium", "Low"))
  )

# Industry & Birthplace
acs_data <- acs_data %>%
  mutate(
    industry_name   = as.character(haven::as_factor(industry_encoded)),
    birthplace_label = haven::as_factor(birthplace)
  )

# Compute Industry Ratio Metrics
ratio_data_detailed <- acs_data %>%
  filter(!is.na(industry_name)) %>%
  group_by(industry_name, citizen_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = citizen_label,
    values_from = count,
    values_fill = 0
  ) %>%
  mutate(
    ratio_nat = `Naturalized Citizen` / `Native Citizen`,
    ratio_non = `Non-Citizen` / `Native Citizen`
  ) %>%
  arrange(desc(ratio_nat + ratio_non))

ratio_long <- ratio_data_detailed %>%
  select(industry_name, ratio_nat, ratio_non) %>%
  pivot_longer(
    cols = starts_with("ratio"),
    names_to = "Group",
    values_to = "Ratio"
  ) %>%
  mutate(
    Group = recode(
      Group,
      ratio_nat = "Naturalized Citizen",
      ratio_non = "Non-Citizen"
    )
  )

# Center Age at the (unweighted) sample mean
mean_age <- mean(acs_data$age, na.rm = TRUE)
acs_data <- acs_data %>%
  mutate(
    age_c    = age - mean_age,
    age_c_sq = age_c^2,
    # Labor supply and the covariates added in the hours revision
    ln_hours    = log(annual_hours),
    marital_lbl = haven::as_factor(marital_status),
    nchlt5_lbl  = haven::as_factor(nchlt5_grouped),
    ftfy_lbl    = factor(ftfy, levels = c(0, 1), labels = c("Not FTFY", "FTFY"))
  )

# Preferred Specification: must stay identical to M6 in 02_Regressions_and_Stats.do
full_fe_model <- feols(
  ln_wage ~ factor(citizen_cat_clean) * factor(sex) +
    age_c + age_c_sq +
    educ_grouped +
    eng_fluency +
    nchild_grouped +
    marital_lbl * factor(sex) +
    nchlt5_lbl +
    factor(rachsing) +
    ln_hours |
    year + industry_encoded + puma_state + birthplace,
  cluster = ~puma_state,
  data = acs_data
)

etable(full_fe_model, digits = 4)
```

# Figures

## Figure 1: Log Wage Transformation & Distributional Symmetry

```{r}
#| fig-width: 8
#| fig-height: 4.5

fig1 <- acs_data %>%
  select(incwage_adj, ln_wage) %>%
  pivot_longer(cols = everything(), names_to = "metric", values_to = "value") %>%
  mutate(metric = ifelse(metric == "incwage_adj", "Raw Annual Wages", "Log Annual Wages")) %>%
  ggplot(aes(x = value, fill = metric)) +
  geom_density(alpha = 0.6, show.legend = FALSE) +
  facet_wrap(~metric, scales = "free") +
  scale_fill_manual(values = c("Log Annual Wages" = "#21618c", "Raw Annual Wages" = "#e74c3c")) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "Distributional Transformation of Wages",
    subtitle = "Demonstrating Skewness Reduction via Natural Log Transformation",
    x = "Value",
    y = "Density"
  ) +
  theme_minimal(base_size = 11)

fig1
```

## Figure 2: Foreign-Born Demographic Diversity by Birthplace

```{r}
#| fig-width: 8
#| fig-height: 5

birthplace_counts <- acs_data %>%
  filter(!birthplace_label %in% c("U.S. & Territories", "Other/Unknown", "N/A")) %>%
  count(birthplace_label) %>%
  arrange(desc(n))

fig2 <- ggplot(birthplace_counts, aes(x = reorder(birthplace_label, -n), y = n)) +
  geom_bar(stat = "identity", fill = "#21618c") +
  geom_text(aes(label = comma(n)), vjust = -0.3, size = 3) +
  labs(
    title = "Demographic Diversity by Birthplace (Foreign-Born Only)",
    x = "Birthplace Region",
    y = "Count of Individuals") +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.1))) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9))

fig2
```

## Figure 3: Industry Concentration by Citizenship Status

```{r}
#| fig-width: 9
#| fig-height: 5

fig3 <- ggplot(ratio_long, aes(x = reorder(industry_name, Ratio), y = Ratio, fill = Group)) +
  geom_col(position = "stack", width = 0.8) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.05))) +
  scale_fill_viridis_d(option = "G", begin = 0.2, end = 0.8) +
  labs(
    title = "Foreign-Born Worker Concentration by Industry and Citizenship Status",
    subtitle = "Ratio of Foreign-Born Workers relative to Native-Born Citizen Workers (ACS 2012-2019)",
    x = "Industry Sector",
    y = "Ratio to Native-Born Citizen Workers (%)",
    fill = "Citizenship Status"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

fig3
```

## Figure 4: Intersectional Income Group Distribution

```{r}
#| fig-width: 10
#| fig-height: 6

educ_income_intersectional <- acs_data %>%
  filter(!is.na(educ_grouped)) %>%
  group_by(educ_grouped, income_group, citizen_label, sex_label) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(educ_grouped, citizen_label, sex_label) %>%
  mutate(percent = count / sum(count) * 100)

fig4 <- ggplot(educ_income_intersectional, aes(x = educ_grouped, y = percent, fill = income_group)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_viridis_d(name = "Income Tertile", option = "D") +
  labs(
    title = "Income Group Distribution by Gender, Citizenship, and Education",
    x = "Educational Attainment Group",
    y = "Percentage within Group"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 8),
    strip.text = element_text(size = 10, face = "bold")
  ) +
  facet_grid(sex_label ~ citizen_label)

fig4
```

## Figure 5: Returns to Education by Citizenship and Gender

```{r}
#| fig-width: 9
#| fig-height: 5.5
#| warning: false

educ_sample <- acs_data %>%
  filter(!is.na(educ_years), !is.na(citizen_sex_lbl), age >= 25)

educ_returns <- educ_sample %>%
  group_by(citizen_sex_lbl, educ_years) %>%
  summarise(
    mean_ln_wage = mean(ln_wage),
    se           = sd(ln_wage) / sqrt(n()),
    n            = n(),
    .groups      = "drop"
  ) %>%
  filter(n >= 30) 

educ_slopes <- educ_sample %>%
  mutate(above_hs = educ_years >= 12) %>%
  group_by(citizen_sex_lbl, above_hs) %>%
  summarise(
    slope = coef(lm(ln_wage ~ educ_years))[2],
    n     = n(),
    .groups = "drop"
  )

print(educ_slopes)

fig5 <- ggplot(educ_returns, aes(x = educ_years, y = mean_ln_wage,
                                 color = citizen_sex_lbl, fill = citizen_sex_lbl)) +
  geom_ribbon(aes(ymin = mean_ln_wage - 1.96 * se, ymax = mean_ln_wage + 1.96 * se),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_color_viridis_d(option = "plasma", name = "Demographic Group") +
  scale_fill_viridis_d(option = "plasma", name = "Demographic Group") +
  scale_x_continuous(breaks = seq(0, 18, by = 2)) +
  labs(
    title = "Returns to Education by Citizenship and Gender",
    subtitle = "Age 25+; wages are flat below 12 years and rise steeply after, for every group",
    x = "Years of Formal Education",
    y = "Mean Log Wage"
  ) +
  theme_minimal(base_size = 11) 

fig5
```

# Saving Images

```{r}
dir.create(here("Output", "figures"), recursive = TRUE, showWarnings = FALSE)

save_fig <- function(filename, plot_obj, w, h) {
  ggsave(
    filename = here("Output", "figures", filename),
    plot = plot_obj,
    width = w, height = h,
    dpi = 300, bg = "white"
  )
}

save_fig("Figure_1_Log_Wages.png",       fig1, 8,  4.5)
save_fig("Figure_2_Birthplace.png",      fig2, 8,  5)
save_fig("Figure_3_Industry.png",        fig3, 9,  5)
save_fig("Figure_4_Income.png",          fig4, 10, 6)
save_fig("Figure_5_Returns_to_Educ.png", fig5, 9,  5.5)

save_pdf <- function(filename, plot_obj, w, h) {
  ggsave(
    filename = here("Paper", filename),
    plot = plot_obj,
    width = w, height = h, bg = "white"
  )
}

save_pdf("fig1_log_wages.pdf",      fig1, 8,  4.5)
save_pdf("fig2_birthplace.pdf",     fig2, 8,  5)
save_pdf("fig3_industry_ratio.pdf", fig3, 9,  5)
save_pdf("fig4_income_dist.pdf",    fig4, 10, 6)
save_pdf("fig5_returns_educ.pdf",   fig5, 9,  5.5)
```
